---
title: "Flavours-of-cocoa"
description: |
  A short description of the post.
author:
  - name: Jonathan Jayes
    url: {interludeone.com}
date: 2022-01-17
output:
  distill::distill_article:
    self_contained: false
    code_folding: true
    highlight: haddock
    highlight_downlit: true
    toc: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F)

```

# Flavours of Cocoa

Welcome to the first in a series of data screencasts where I attempt to show you how great the R language is.

This post follows along with the data screencast and includes the code methodically.

It begins by reading in the data, then tidying it, analysing it, making some visualizations and then performing some predictive modelling.

## What's the deal with beans?

I'm signed up to a fantastic newsletter called ["Data is Plural"](https://www.data-is-plural.com/) curated by journalist Jeremy Singer-Vine. Truly, it is a treasure trove of interesting public datasets from all over the internet. You can sign up to the newsletter [here](https://buttondown.email/data-is-plural) if this sounds up your alley. 

This week it included a link to a fun selection of chocolate bar reviews, which Jeremy described as:

<blockquote>

Chocolate bar reviews. The [Manhattan Chocolate Society](http://flavorsofcacao.com/mcs_index.html)’s [Brady Brelinski](http://flavorsofcacao.com/contact.html) has reviewed 2,500+ bars of craft chocolate since 2006, and compiles his findings into a [copy-paste-able](http://flavorsofcacao.com/chocolate_database.html) table that lists each bar’s manufacturer, bean origin, percent cocoa, ingredients, review notes, and numerical rating.

</blockquote>

A live link to the database is shown below.

```{r}
knitr::include_url("http://flavorsofcacao.com/chocolate_database.html")
```


## Ingest the data

I've copied and pasted the data into an Excel spreadsheet, accessible on my [Github]() if you want to download it an analyse it yourself.

<aside>

The underlying data is served in a Javascript container on the website rather than vanilla HTML. This makes it a little bit more difficult to scrape with a package like `rvest`, for example. Hence the Excel spreadsheet.

</aside>
```{r}

```

